Transition path sampling and the calculation of free energies of enzymatic reactions

### 2022-03-31 tex/manuscript.tex
added some more stuff

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Molecular dynamics (MD) simulations are being 
used to calculate quantitative estimations of experimentally measurable 
kinetic and thermodynamic parameters of enzyme catalyzed reactions. 
\cite{Karplus02NatStructMolBiol9p646,
Schramm98AnnuRevBiochem67p693,Zhang05AccChemRes38p379,
Schramm11AnnuRevBiochem80p703,Schramm18ChemRev118p11194} 
This has helped in proposing mechanisms and explanations for 
enzyme activity, high throughput screening of millions of inhibitor 
molecues in drug discovery,

This is quite useful in explaining 
the efficiency and rate of such reactions, developing 
inhibitors for various enzymes in pharmacology, de novo design of 
enzymatic activity.
MD simulations model chemical reactions by taking the system from a reactant state 
to a product state via a high energy transition state.
To overcome these energy barriers and 
to obtain ergodic sampling several methods have been developed over the years such as 
umbrella sampling, \cite{Kastner11WileyInterdiscipRevComputMolSci1p932} 
metadynamics, \cite{Barducci11WileyInterdiscipRevComputMolSci1p826}
steered MD, \cite{Park04JChemPhys120p5946}
milestoning, etc. These methods rely on applying bias
potentials for the system to overcome energy barriers and hence require the knowledge of 
the reaction coordinates.  
Transition path sampling \cite{Bolhuis02AnnRevPhysChem53p291,Dellago98JChemPhys108p1964}
(TPS) is an enhanced sampling method for 
rare events that transition a system between two stable states often 
referred to as reactants and products. 6

@@

One example of the TPS based free energies calculations on real protein systems \cite{Silveira21JPhysChemB125p2018}

Human Mat2a enzyme and plasmodium vivax adenosine deaminase \cite{Luo07JAmChemSoc129p8008,Ho09Biochemistry48p9618} 
\section{Methods}
\subsection{Transition path sampling}
%Transition path sampling is an importance sampling in the trajectory space.
Consider the molecular dynamics simulation of a molecular system consisting 
of $N$ atoms. Starting from the initial positions and momenta of each atom at time $t_0$,
the time evolution is carried out by finding the position ($\textbf{q}$) and 
momenta ($\textbf{p}$) of each atom at regular intervals of time $t_i = t_0 + i\Delta t$  
($i = 0,1,2,\ldots$). At each of the time slices, the positions and momenta can be collectively represented as
the set $z = \{\textbf{q},\textbf{p}\}$. If the MD simulation is run for a total 
time of $\mathcal{T}$ the number of time slices is given by $L = \mathcal{T}/\Delta t +1$ and for this sequence of 
times the state of the system can be represented as 
\begin{equation}
z(\mathcal{T}) = \{z_0, z_{\Delta t}, z_{2\Delta t},\ldots,z_{\mathcal{T}}\}
\end{equation}
The probability to obtain a particular sequence of states is determined by the initial 
conditions and the type of dynamics used for the time evolution. If the time evolution is 
Markovian, the probability to go from $z_{i\Delta t}$ to $z_{(i+1)\Delta t}$ depends only 
on $z_{i\Delta t}$ and not on conditions before the time $i\Delta t$ the total path probablility can 
be expressed as the product of individual probabilities $p(i\rightarrow j)$ as 
\begin{equation}
\mathcal{P}[z(\mathcal{T})] = \rho(z_0)\Pi_{i=0}^{\mathcal{T}/\Delta t-1} p(z_{i\Delta t}\rightarrow z_{(i+1)\Delta t}),
\end{equation}
where $\rho(z_0)$ is the equilibrium distribution of the initial conditions and for a canonical ensemble this can be 
expressed as 
\begin{equation}
\rho(z_0) = \exp(-\beta H(z_0))/Z
\end{equation}
where 
\begin{equation}
Z = \int dz \exp(-\beta H(z)) 
\end{equation}
is the canonical partition function.

@@

\subsection{Equilibrium distribution of order parameters}
%Monte carlo methods in trajectory space the probability of a path is 
%\begin{equation}
%P(x_A)
%\end{equation}
The free energy as a function of the order parameter $\zeta$ is defined
as 
\begin{equation}
\text{A}(\zeta) = -k_{\text{B}}T\text{ln}(P(\zeta)) + C, \label{eqn:fenergy}
\end{equation}
where $P(\zeta)$ is the probability distribution of the reaction coordinate
$\zeta$ and $C$ is an arbitrary constant. At equilibrium the probability distribution of the 
reaction coordinate 
can be obtained from the distribution function for the phase space $\rho(\textbf{q})$ as
\begin{equation}
P(\zeta) = \int d\textbf{V} \rho(\textbf{q})\delta\left[\zeta-\tilde{\zeta}(\textbf{q})\right].
\end{equation}
The integration is over the entire phase space. In practice the calculation of this distribution function
often proceeds using histogram based methods where the reaction coordinate is 
divided into bins and the frequency of the occurence of the reaction coordinate within a particular window
during the course of molecular dynamics simulations is used to calculate the probability distribution. 
Since some values of the reaction coordinates particulary close to the transition state are 
rarely sampled in a conventional MD simulations, enhanced sampling techniques such as the umbrella sampling
are necessary to sample the low probability regions. TPS is designed to sample these regions in phase space 
without the application of any external bias, but the trajectories sampled with TPS are only reactive and hence
they are not distributed according to the equilibrium distribution function. 
The restriction that the pathways start from the reactant state and end in the product state can be relaxed
to obtain an ensemble of trajectories that will be distributed according to the equilibrium distribution.

\subsection{The algorithm}
We follow the steps proposed by Radhakrishnan et al. \cite{Radhakrishnan04JChemPhys121p2436} 
to sample TPS trajectories within windows of the order parameter. 
\begin{itemize}
  \item An appropriate order parameter is chosen for the reaction of interest 
  that can distinguish the reactant and product states sufficiently well and the TPS ensemble is harvested 
  which consists only of reactive trajectories.
\item Divide the order parameter into windows $\{\zeta_i\}$ within which the 
order parameters are given by 
$\zeta_{i}^{min} < \zeta < \zeta_{i}^{max}$.
\item Choose a reactive trajectory $\tilde{z}(\mathcal{T})$ from the TPS ensemble as 
a guiding trajectory for the window sampling. Pick a time slice from this trajectory 
$\tilde{z}_{j\Delta t}$ such that the order parameter calculated for this time slice
$\tilde{\zeta}$ satisfies $\zeta_{i}^{min} < \tilde{\zeta} < \zeta_{i}^{max}$.
\item Using the shooting and shifting algorithm familiar from TPS simulations, harvest 
dynamics trajectories starting from the time slice $\tilde{z}_{j\Delta t}$ of the trajectory 
  $\tilde{z}(\mathcal{T})$ and accept the new trajectory if at any time slice the calculated 
  order parameter $\hat{\zeta}$ satisfies $\zeta_{i}^{min} < \hat{\zeta} < \zeta_{i}^{max}$.
  \item Calculate the probability distribution ($P_i(\zeta)$) within the window $\zeta_{i}^{min} 
  < \zeta < \zeta_{i}^{max}$ by constructing histograms obtained from the ensemble of 
  accepted trajectories for the window $i$. 
  \item Combine the probability distributions in successive windows by adjusting the constants $C$
  such that the free energy $A$ from Eq. \ref{eqn:fenergy} is continuous. 
\end{itemize}

\section{Results and discussion}
\subsection{QMMM simulations}
CHARMM force field \cite{Brooks09JComputChem30p1545}

\section{Adenosine deaminase reaction coordinates and free energy}
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.22]{./figures/ada-protein.png}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Molecular dynamics (MD) simulations are increasingly being
used to calculate quantitative estimations of experimentally measurable 
kinetic and thermodynamic parameters of enzyme catalyzed reactions. 
\cite{Karplus02NatStructMolBiol9p646,
Schramm98AnnuRevBiochem67p693,Zhang05AccChemRes38p379,
Schramm11AnnuRevBiochem80p703,Schramm18ChemRev118p11194} 
This has helped in proposing mechanisms and explanations for 
enzyme activity, fast high throughput screening of thousands of inhibitor 
molecues in drug lead discovery, \cite{Jorgensen09AccChemRes42p724,Sliwoski14PharmacolRev66p334} 
directed evolution for designing enzymes with enhanced catalytic 
activities, \cite{Thyme09Nature461p1300,Bloom09PNAS106p9995,Schafer19JAmChemSoc141p10431} etc.    
Straightforward MD simulations are not employed in these cases

Ergodic sampling of the phase space within straighforward MD simulations 
is often difficult to achieve with the presence of energy barriers. 
To overcome such barriers enhanced sampling techniques 
have been developed over the years such as 
umbrella sampling, \cite{Kastner11WileyInterdiscipRevComputMolSci1p932} 
metadynamics, \cite{Barducci11WileyInterdiscipRevComputMolSci1p826}
steered MD, \cite{Park04JChemPhys120p5946}
milestoning, \cite{Faradjian04JChemPhys120p10880} etc. These methods rely on applying bias
potentials for the system to overcome energy barriers and hence 
require the knowledge of the reaction coordinates.  
Transition path sampling \cite{Bolhuis02AnnRevPhysChem53p291,Dellago98JChemPhys108p1964}
(TPS) is an enhanced sampling method for 
rare events that transition a system between two stable states often 
referred to as reactants and products.

@@

In this work we will focus on the implementation and application of the 
free energy based calculations within the TPS method on two enzymatic reactions: 
One example of the TPS based free energies calculations on real 
protein systems \cite{Silveira21JPhysChemB125p2018} Human Mat2a enzyme and 
plasmodium vivax adenosine deaminase \cite{Luo07JAmChemSoc129p8008,Ho09Biochemistry48p9618}
%------------------------------------------------------------------------------ 
\section{Methods}
%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------ 
%Transition path sampling is an importance sampling in the trajectory space.
Consider the molecular dynamics simulation of a molecular system consisting 
of $N$ atoms. Starting from the initial positions and momenta of each atom at time $t_0$,
the time evolution is carried out by finding the position ($\textbf{q}$) and 
momenta ($\textbf{p}$) of each atom at regular intervals of time $t_i = t_0 + i\Delta t$  
($i = 0,1,2,\ldots$). At each of the time slices, the positions and momenta can be collectively represented as
the set $z = \{\textbf{q},\textbf{p}\}$. If the MD simulation is run for a total 
time of $\mathcal{T}$ the number of time slices is given by $L = \mathcal{T}/\Delta t +1$ and for this sequence of 
times the state of the system can be represented as 
\begin{equation}
z(\mathcal{T}) = \{z_0, z_{\Delta t}, z_{2\Delta t},\ldots,z_{\mathcal{T}}\}
\end{equation}
The probability to obtain a particular sequence of states is determined by the initial 
conditions and the type of dynamics used for the time evolution. If the time evolution is 
Markovian, the probability to go from $z_{i\Delta t}$ to $z_{(i+1)\Delta t}$ depends only 
on $z_{i\Delta t}$ and not on conditions before the time $i\Delta t$ the total path probablility can 
be expressed as the product of individual probabilities $p(i\rightarrow j)$ as 
\begin{equation}
\mathcal{P}[z(\mathcal{T})] = \rho(z_0)\Pi_{i=0}^{\mathcal{T}/\Delta t-1} p(z_{i\Delta t}\rightarrow z_{(i+1)\Delta t}),
\end{equation}
where $\rho(z_0)$ is the equilibrium distribution of the initial conditions and for a canonical ensemble this can be 
expressed as 
\begin{equation}
\rho(z_0) = \exp(-\beta H(z_0))/Z
\end{equation}
where 
\begin{equation}
Z = \int dz \exp(-\beta H(z)) 
\end{equation}
is the canonical partition function.

@@

\subsection{Equilibrium distribution of order parameters}
%Monte carlo methods in trajectory space the probability of a path is 
%\begin{equation}
%P(x_A)
%\end{equation}
The free energy as a function of the order parameter $\zeta$ is defined
as 
\begin{equation}
\text{A}(\zeta) = -k_{\text{B}}T\text{ln}(P(\zeta)) + C, \label{eqn:fenergy}
\end{equation}
where $P(\zeta)$ is the probability distribution of the reaction coordinate
$\zeta$ and $C$ is an arbitrary constant. At equilibrium the probability distribution of the 
reaction coordinate 
can be obtained from the distribution function for the phase space $\rho(\textbf{q})$ as
\begin{equation}
P(\zeta) = \int d\textbf{V} \rho(\textbf{q})\delta\left[\zeta-\tilde{\zeta}(\textbf{q})\right].
\end{equation}
The integration is over the entire phase space. In practice the calculation of this distribution function
often proceeds using histogram based methods where the reaction coordinate is 
divided into bins and the frequency of the occurence of the reaction coordinate within a particular window
during the course of molecular dynamics simulations is used to calculate the probability distribution. 
Since some values of the reaction coordinates particulary close to the transition state are 
rarely sampled in a conventional MD simulations, enhanced sampling techniques such as the umbrella sampling
are necessary to sample the low probability regions. TPS is designed to sample these regions in phase space 
without the application of any external bias, but the trajectories sampled with TPS are only reactive and hence
they are not distributed according to the equilibrium distribution function. 
The restriction that the pathways start from the reactant state and end in the product state can be relaxed
to obtain an ensemble of trajectories that will be distributed according to the equilibrium distribution. 
%------------------------------------------------------------------------------ 
\subsection{The algorithm}
%------------------------------------------------------------------------------ 
We follow the steps proposed by Radhakrishnan et al. \cite{Radhakrishnan04JChemPhys121p2436} 
to sample TPS trajectories within windows of the order parameter. 
\begin{itemize}
  \item An appropriate order parameter is chosen for the reaction of interest 
  that can distinguish the reactant and product states sufficiently well and the TPS ensemble is harvested 
  which consists only of reactive trajectories.
\item Divide the order parameter into windows $\{\zeta_i\}$ within which the 
order parameters are given by 
$\zeta_{i}^{min} < \zeta < \zeta_{i}^{max}$.
\item Choose a reactive trajectory $\tilde{z}(\mathcal{T})$ from the TPS ensemble as 
a guiding trajectory for the window sampling. Pick a time slice from this trajectory 
$\tilde{z}_{j\Delta t}$ such that the order parameter calculated for this time slice
$\tilde{\zeta}$ satisfies $\zeta_{i}^{min} < \tilde{\zeta} < \zeta_{i}^{max}$.
\item Using the shooting and shifting algorithm familiar from TPS simulations, harvest 
dynamics trajectories starting from the time slice $\tilde{z}_{j\Delta t}$ of the trajectory 
  $\tilde{z}(\mathcal{T})$ and accept the new trajectory if at any time slice the calculated 
  order parameter $\hat{\zeta}$ satisfies $\zeta_{i}^{min} < \hat{\zeta} < \zeta_{i}^{max}$.
  \item Calculate the probability distribution ($P_i(\zeta)$) within the window $\zeta_{i}^{min} 
  < \zeta < \zeta_{i}^{max}$ by constructing histograms obtained from the ensemble of 
  accepted trajectories for the window $i$. 
  \item Combine the probability distributions in successive windows by adjusting the constants $C$
  such that the free energy $A$ from Eq. \ref{eqn:fenergy} is continuous. 
\end{itemize}   
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{QMMM simulations}
%------------------------------------------------------------------------------ 
CHARMM force field \cite{Brooks09JComputChem30p1545}

%------------------------------------------------------------------------------ 
\subsection{Adenosine deaminase reaction coordinates and free energy}
%------------------------------------------------------------------------------ 
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.22]{./figures/ada-protein.png}
\end{figure}
